Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-05-12 endofquarterpaper.qmd
hunter-gatherer edits

Variation in diversity of sign types among sites may be representative of variation in social arrangement or social context. Despite the tendency to conceive of hunter-gatherers as monolithic, previous studies have established a potential for variation in hunter-gatherer social arrangements [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017]. @Bloch_1977 drew upon ethnographic studies of hunter-gatherers to theorize a correlation between the amounts of social structure, instituted hierarchy, and ritual communication within a society. @Conkey_1985 built upon that theorized correlation to argue that artwork and other forms of material culture can serve as a form of ritual communication, and that an increase in diversity within those material cultures may suggest more elaborate or complex social contexts. If the signs served functions of ritual communication, then changes in sign diversity from site to site, particularly the larger grouping structure of broader versus restricted signs, may be representative of two different social arrangement types, such as, for example, a more hierarchical structure versus a less hierarchical structure.

@@

One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites. The aggregation/dispersion pattern is first defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons. Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994]. It is therefore possible that sites with multiple connections between groups [@fig-neighbor-net], such as La Ferrassie in the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.

```{r}

@@

# Conclusion

This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996]. @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our results show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted. We attempted to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023] by dividing the data up into smaller time periods. Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].

>>>>

@

Variation in diversity of sign types among sites may indicate variation in social structure. Despite the tendency to conceive of hunter-gatherers as monolithic, previous studies have established a potential for variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006]. Additional research has illustrated how ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978;@Tilley_1982]. Therefore, it is possible that greater diversity in geometric signs and other forms of material culture can serve as evidence of wider variety of formalized social roles [@Conkey_1985]. This is especially relevant in regards to our larger grouping structure of broader versus restricted signs, as it may indicate two broad types of social structure, with the broader groups having a more formalized social structure, and the restricted groups having a less formalized social structure.

@@

One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites. The aggregation/dispersion pattern is first defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons. Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].

It is therefore possible that some of the sites shown in @fig-div-plot which have a more diverse range of sign types than predicted by sample size, such as Les Cottés in the Proto-Aurignacian, Solutré in the Early Aurignacaibn, and Bockstein-Törle and Vogelherd in the Evolved Aurignacian acted as aggregation sites, or served a similar purpose.

It is therefore possible that sites with multiple connections between groups [@fig-neighbor-net], such as La Ferrassie in the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.

```{r}
#| label: fig-div-plot
#| fig-cap: "Plots of expected versus actual diversity of signs for each site by time period. The blue line is the expected Heterogeneity value under the null model (simple increase in sample size means simple increase in diversity of sign types. The upper red line is the upper 95% confidence interval, if a site is near or above that line, it has a much higher diversity of signs than expected under the null model.The lower red line is the lower 95% confidence interval, if a site is near or below that line, it has a much lower diversity of signs than expected under the null model."

@@

# Conclusion
This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996]. @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our results show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted. We attempted to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023] by dividing the data up into smaller time periods. Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].
